\abstract{%
\textbf{Problem.} Cell-free DNA (cfDNA) whole-genome sequencing has become the dominant non-invasive substrate for non-invasive prenatal testing, minimal residual disease monitoring, and early cancer detection, but methylation information remains expensive to access because bisulfite conversion is destructive and biased. \citet{liu2024finaleme} closed part of this gap with FinaleMe, a per-sample two-state hidden Markov model that recovers methylation from fragmentomic features alone. Its binary latent state, point-estimate Baum-Welch fit, per-sample training, susceptibility to prior leakage, lower-information CpG regimes, and binarized downstream tissue deconvolution leave room for a fully probabilistic successor.

\smallskip\textbf{Method.} We present \textbf{HAVI-Methyl} (Hierarchical Amortized Variational Inference for Methylation), a hierarchical Bayesian framework for methylation prediction from cfDNA fragmentomics. The full proposed model uses a population locus prior $\beta^{\mathrm{pop}}_\ell$, a per-sample shift $\delta_s$, and a per-(sample, locus) latent $\beta_{s,\ell}$ on the logit scale, with direct-count, pseudo-count, end-motif, and coverage likelihoods chosen according to the observation regime. Inference is designed as stochastic variational inference with approximate moment-matched natural-gradient updates on the population layer and an amortized normalizing-flow variational posterior $q_\phi(\eta_{s,\ell}\mid F_{s,\ell})$ over per-locus fragment bags, encoded by a permutation-invariant Set Transformer and optional frozen long-range DNA embeddings. Prior-leakage control is handled by a variational information-bottleneck penalty, mQTL anchor losses, and counterfactual augmentation. Calibrated prediction sets come from posterior-predictive checks plus a split-conformal wrapper, and the proposed tissue-of-origin module uses posterior methylation uncertainty rather than binarized point estimates.

\smallskip\textbf{Results.} On the real Liu 2024 paired cfDNA WGS / WGBS panel ($S=77$ patients, $L=782$ high-variance CpGs, manifest-paired via the published Supplementary Table~1), the full HAVI-Methyl torch loop (Set Transformer encoder + Gaussian posterior head + Beta-Binomial reconstruction + Robbins-Monro population shifts) trained 200 iterations on an A10G GPU achieves Pearson $r = 0.455$ versus $r = 0.078$ for a FinaleMe-style HMM baseline ($\sim 5.8\times$), AUC $0.74$ versus $0.56$, and reduces credible-interval ECE from $0.47$ to $0.32$. Stratifying by WGBS depth, HAVI-Methyl is the only method with positive correlation in every stratum; FinaleMe is anti-correlated with truth ($r=-0.05$) in the multi-read interior stratum. On the published Loyfer/UXM\_deconv U25 panel ($36$ cell types $\times$ $900$ marker blocks), the variance-weighted Dirichlet head wins LOO RMSE at every one of the $36$ cell types (overall $0.017$ vs lstsq $0.028$ vs FinaleMe-binarized $0.038$). On a synthetic multi-seed recovery benchmark, HAVI-Methyl beats the FinaleMe-style HMM at every fragment-bag coverage with non-overlapping $90\%$ bootstrap intervals at $0.1\times$, $1\times$, and $5\times$. A synthetic prior-leakage stress test reduces prior attribution from $5.93\%$ to $0.62\%$ via VIB and to $0.037\%$ with mQTL anchors.

\smallskip\textbf{Contributions.} (i)~A hierarchical amortized variational inference framework for cfDNA methylation; (ii)~a Set Transformer + Conditional NSF flow + Beta-Binomial torch SVI training stack with measured end-to-end behaviour on real data; (iii)~formal propositions covering lower-bound validity, conditional instrumental-anchor identifiability, VIB-bounded prior leakage, conformal marginal coverage, surrogate SVI convergence, hierarchical-pooling variance reduction, and information-theoretic recoverability limits; (iv)~a chromatin-aware cfDNA fragmentation simulator; (v)~real-data Liu 2024 paired methylation benchmark with manifest-based donor pairing; (vi)~Loyfer U25 tissue-of-origin LOO; (vii)~$N{=}20$-seed bootstrap intervals.
}
