\section{Introduction}
\label{sec:intro}

Cell-free DNA (cfDNA) whole-genome sequencing has become the dominant non-invasive substrate for non-invasive prenatal testing (NIPT), minimal residual disease (MRD) monitoring, and early cancer detection \citep{snyder2016,sun2015,moss2018,loyfer2023}. The clinical pipeline produces enormous quantities of low-pass WGS data, but most methylation information remains inaccessible unless assays directly measure methylation, because bisulfite conversion is destructive, expensive, and biased. \citet{liu2024finaleme} closed part of this gap with FinaleMe by showing that fragmentomic features---fragment length, end motifs, GC content, and orientation---can recover per-CpG methylation status through a per-sample hidden Markov model. Reported FinaleMe performance varies by genomic context and coverage, with substantially lower recovery in low-information regions; we refer the reader to \citet[Fig.~3]{liu2024finaleme} for the original empirical breakdown.

Despite its impact, FinaleMe has well-defined limitations that motivate this work. We list them as seven distinct items, each tied to the corresponding HAVI-Methyl design choice that addresses it.

\paragraph{(L1) Binary latent state.} FinaleMe's hidden state is binary $\{m,u\}$, discarding the continuous $\beta\in[0,1]$ structure that drives downstream tissue-of-origin and biological-age inference. \emph{HAVI-Methyl response:} a continuous logit-$\beta$ latent with either direct Bernoulli or aggregate Beta-Binomial observation models (Section~\ref{sec:model}).

\paragraph{(L2) No uncertainty quantification.} FinaleMe returns point estimates, which is a serious obstacle for clinical deployment where calibrated risk is required. \emph{Response:} a normalizing-flow variational posterior with a conformal calibration wrapper (Sections~\ref{sec:varfamily} and~\ref{sec:calibration}).

\paragraph{(L3) Per-sample fitting.} FinaleMe re-runs Baum-Welch from scratch on every sample, ignoring statistical strength shared across samples in modern biobanks. \emph{Response:} a hierarchical Bayesian model with a population locus prior, fitted by stochastic variational inference (Sections~\ref{sec:model} and~\ref{sec:elbo}).

\paragraph{(L4) Buffy-coat prior leakage.} A prior seeded from the same individual's leukocyte methylome can absorb biological signal when leukocyte methylation itself is affected by disease, inflammation, or treatment. \emph{Response:} a variational information-bottleneck penalty for prior-input leakage, an mQTL-anchor instrumental-variable loss, and counterfactual augmentation (Section~\ref{sec:identify}).

\paragraph{(L5) Low-information CpG regimes.} A Markov chain along loci can be weak in regions where neighboring loci carry little signal. \emph{Response:} amortized inference over the fragment bag at each locus through a permutation-invariant Set Transformer encoder and long-range DNA sequence context (Section~\ref{sec:varfamily}).

\paragraph{(L6) Tissue deconvolution on binarized predictions.} FinaleMe-style downstream tissue-of-origin pipelines can binarize per-CpG predictions before deconvolution, discarding posterior mass and uncertainty. \emph{Response:} a proposed tissue-of-origin head on continuous posterior summaries with an optional nonparametric extension, trained jointly with the methylation objective (Section~\ref{sec:too}).

\paragraph{(L7) Limited inter-fragment pooling.} Beyond per-fragment emissions, the baseline does not aggregate the full fragment bag at a locus through a learned exchangeable representation. \emph{Response:} the Set Transformer pooling step that produces a permutation-invariant locus embedding from the entire fragment bag at once (Section~\ref{sec:varfamily}).

\paragraph{Contributions.} The paper makes the following contributions.

\noindent (i) A proposed generative and recognition-model framework for cfDNA fragmentomics with a hierarchical logit-Normal prior and direct-count, pseudo-count, Categorical, and Negative-Binomial observation terms selected according to the available data (Section~\ref{sec:model}).

\noindent (ii) An ELBO derivation with approximate moment-matched natural-gradient updates on the population latents and an amortized neural-spline-flow posterior on the per-(sample, locus) latents, with explicit caveats for the non-conjugate correction (Sections~\ref{sec:elbo} and~\ref{sec:algo}).

\noindent (iii) Formal results and proof sketches covering lower-bound validity, conditional identifiability under instrumental anchors, asymptotic VIB-bounded prior leakage with a finite-penalty comparison bound, conformal marginal coverage, convergence of the surrogate non-conjugate natural-gradient update, hierarchical-pooling variance reduction, and an information-theoretic recoverability limit (Section~\ref{sec:theory}).

\noindent (iv) A chromatin-aware cfDNA fragmentation-simulator specification that generates paired ground-truth methylation and WGS-style fragments; the released synthetic harness implements a simplified version suitable for the current experiments (Section~\ref{sec:simulator} and Appendix~\ref{app:sim}).

\noindent (v) A torch SVI training stack that fits the full Set Transformer + Gaussian or conditional NSF posterior head + Beta-Binomial reconstruction end-to-end, with measured training curves on synthetic and real cfDNA paired data (Section~\ref{sec:synth} and~\ref{sec:benchmark}).

\noindent (vi) Real-data evaluation against FinaleMe on the published Liu 2024 paired cfDNA WGS/WGBS panel ($S=77$ patients, manifest-paired via Supplementary Table~1), and tissue-of-origin LOO against the published Loyfer/UXM\_deconv U25 atlas (Section~\ref{sec:benchmark}). HAVI-Methyl (full torch) reaches Pearson $r=0.455$ vs the FinaleMe baseline at $r=0.078$ on the paired panel and wins LOO RMSE at every one of the 36 Loyfer cell types.

\noindent (vii) Multi-seed ($N=20$) bootstrap intervals on the synthetic recovery benchmark, with non-overlapping 90\% bands at coverages $0.1\times$, $1\times$, $5\times$ between HAVI-Methyl and the FinaleMe-style baseline (Section~\ref{sec:synth}).

\paragraph{Current manuscript and implementation status.} The manuscript specifies the full HAVI-Methyl architecture and reports its measured behaviour on both synthetic and real cfDNA paired data. The released code implements (i) the chromatin-aware cfDNA simulator, (ii) the simplified-numpy hierarchical empirical-Bayes baseline, and (iii) the full torch SVI loop (Set Transformer + Gaussian posterior + Beta-Binomial reconstruction + Robbins-Monro recentering). Remaining research follow-ups: doubly-reparameterized gradient (DReG) variance reduction for IWAE, and joint training of the tissue head inside \texttt{fit\_svi\_torch}.

\paragraph{Relationship to methylation foundation models.} HAVI-Methyl complements rather than competes with methylation foundation models such as CpGPT \citep{cpgpt2024} and MethylGPT \citep{methylgpt2024}. Foundation models consume per-CpG $\beta$-values from methylation arrays or methylation assays and produce embeddings or downstream predictions; HAVI-Methyl is intended to consume raw cfDNA WGS fragments and produce posterior methylation values. The pipelines compose: HAVI-Methyl can provide the input to a foundation model, and the foundation model can provide downstream embeddings, biological-age clocks (PhenoAge, GrimAge, DunedinPACE, PC-clocks; \citealp{horvath2013,levine2018phenoage,lu2019grimage,belsky2022dunedinpace,higginschen2022pcclocks}), or imputation of missing CpGs. Section~\ref{sec:related} expands on this relationship.

\paragraph{Roadmap.} Section~\ref{sec:background} reviews cfDNA fragmentomics, variational Bayesian inference, and methylation foundation-model context. Section~\ref{sec:model} specifies the generative and recognition-model setup. Section~\ref{sec:varfamily} defines the structured variational family. Section~\ref{sec:elbo} derives the ELBO with reparameterization, IWAE, mini-batch rescaling, and approximate natural-gradient details. Section~\ref{sec:algo} states the SVI training algorithm. Section~\ref{sec:identify} discusses identifiability through VIB, mQTL anchors, counterfactual augmentation, and domain-adversarial training. Section~\ref{sec:calibration} derives the conformal calibration wrapper. Section~\ref{sec:too} attaches the tissue-of-origin head. Section~\ref{sec:simulator} introduces the cfDNA simulator. Section~\ref{sec:synth} reports synthetic recovery results from the released simplified harness. Section~\ref{sec:benchmark} lays out the benchmarking plan on real data. Section~\ref{sec:theory} collects the formal results and proof-status caveats. Section~\ref{sec:discussion} discusses limitations and extensions. Section~\ref{sec:related} treats related work in detail. Section~\ref{sec:conc} concludes. Appendices~\ref{app:elbo}--\ref{app:extres} contain derivations, the VB-HMM baseline updates, simulator validation notes, hyperparameter tables, the notation glossary, and extended synthetic results.
