\section{Benchmarking Plan on Real Data}
\label{sec:benchmark}

\subsection{Datasets}
We propose to evaluate on the datasets summarized in Table~\ref{tab:datasets}. The paired WGS/WGBS plasma samples of \citet{liu2024finaleme} provide the head-to-head FinaleMe comparison; the Sun-2015, Moss-2018, and Loyfer-2023 resources provide reference panels or tissue-of-origin context. Dataset accessions, sample counts, modalities, and ground-truth availability should be externally verified before final submission. WGBS, methylation arrays, or orthogonal cytometry must be held out according to the protocol below and not reused as atlas initialization for the same test fold.

\begin{table}[htbp]
\centering\small
\caption{Public or controlled-access datasets in the planned benchmark. Counts and access details are to be verified before final submission.}
\label{tab:datasets}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllll}
\toprule
Dataset & $S$ / scope & Mean coverage / modality & Ground truth role & Access \\
\midrule
\citet{liu2024finaleme} & $80$ & paired WGS/WGBS & methylation prediction & EGA \\
\citet{sun2015}        & $32$ & WGBS & reference / validation & SRA \\
\citet{moss2018}        & $25$ & methylation array & ToO reference / validation & GEO \\
\citet{loyfer2023}      & $39$ cell types from $205$ samples & WGBS atlas & ToO reference & GSE186458 \\
\bottomrule
\end{tabular}%
}
\end{table}

\subsection{Metrics}
The benchmark should report per-CpG Pearson and Spearman correlation; ROC-AUC for thresholded binary methylation at $\beta=0.5$; interval expected calibration error computed by binning nominal posterior or conformal levels and comparing empirical coverage; intraclass correlation coefficient under the ICC(2,1) two-way random-effects single-rater convention of \citet{higginschen2022pcclocks}; DMR detection F1 against WGBS-derived DMRs called at $\Delta\beta\ge 0.25$ and Benjamini-Hochberg $q\le 0.05$; and tissue-fraction RMSE against orthogonal flow cytometry or synthetic mixture truth where available. Calibration should also be reported by coverage decile, low-information/CpG-density stratum, chromatin state, GC, mappability, and worst stratum.

\subsection{Ablations}
We separate external baselines from nested HAVI-Methyl ablations. The external baselines are FinaleMe, DeepCpG-style point prediction, and elastic-net regression over pooled fragment features for methylation; CelFiE/CelFEER-style approaches for deconvolution; and foundation-model downstream evaluation as a consumer of inferred methylation. Nested ablations begin from the HAVI-Methyl scaffold and remove or add components one at a time, as summarized in Table~\ref{tab:ablations}.

\begin{table}[htbp]
\centering\small
\caption{Planned nested HAVI-Methyl ablations. External baselines are evaluated separately and are not treated as rows in the same nested sequence.}
\label{tab:ablations}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lccccc}
\toprule
Configuration & BetaBin & Hierarchy & Flow & VIB/mQTL & Conformal \\
\midrule
A0. Feature regression scaffold & & & & & \\
A1. + Beta-Binomial pseudo-likelihood & \checkmark & & & & \\
A2. + hierarchical SVI & \checkmark & \checkmark & & & \\
A3. + amortized flow posterior & \checkmark & \checkmark & \checkmark & & \\
A4. + leakage-control terms & \checkmark & \checkmark & \checkmark & \checkmark & \\
A5. + conformal wrapper (full) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
\bottomrule
\end{tabular}%
}
\end{table}

\subsection{Baselines}
We compare against (i) FinaleMe \citep{liu2024finaleme} as the primary methylation-prediction baseline, (ii) DeepCpG \citep{angermueller2017deepcpg} or an analogous deep-learning point-estimator baseline when inputs can be aligned fairly, (iii) elastic-net regression of $\beta$ on pooled fragment features as a non-deep regression baseline, and (iv) CelFiE/CelFEER-style methods \citep{caggiano2021celfie,keukeleire2023celfeer} for tissue-of-origin deconvolution. MethylGPT \citep{methylgpt2024} is a downstream foundation-model evaluation on inferred methylation rather than a direct methylation-prediction baseline, because it consumes $\beta$ values rather than raw WGS fragments.

\subsection{Evaluation protocol}
The planned protocol uses leave-one-sample-out or grouped cross-validation for sample-level generalization and a chromosome holdout for locus-level generalization. These should be run as distinct leakage checks rather than conflated into one split: sample folds test new-individual prediction, while held-out chromosomes test interpolation across genomic context. DMRs are defined at $\Delta\beta\ge 0.25$ and Benjamini-Hochberg-corrected $q\le 0.05$ in WGBS-derived truth, then matched to predicted DMRs under the same thresholds. All metrics should be reported with bootstrap confidence intervals over samples or genomic blocks, with the resampling unit stated explicitly.

\subsection{Leakage and atlas controls}
Empirical-Bayes initialization from public atlases is allowed only when the atlas samples and loci do not leak held-out targets. The coding agent should maintain a provenance table for every atlas, reference panel, and prior input; exclude test individuals from any initialization; and report a sensitivity analysis comparing random, atlas, and train-only empirical-Bayes initialization.

\subsection{Compute budget on wafer-scale accelerator}
The compute budget is a planning estimate, not a measured throughput claim. The full encoder is expected to have approximately $1.0\,\text{M}$ trainable neural parameters, while the non-amortized variational parameters scale with loci and samples. At $|B_s|=8$, $|B_\ell|=4096$, and mean fragments per locus $\bar n=20$, the dominant costs are ISAB attention and flow evaluation. The final implementation should benchmark sustained throughput, data-loading time, DNA-encoder caching, memory use, and wall-clock time on the actual hardware before publication. Appendix~\ref{app:arch} reports the current planning arithmetic and should be regenerated after the implementation is complete.

% ============================================================
