\section{Theoretical Results}
\label{sec:theory}

This section collects all formal results in one place. Propositions~\ref{prop:elbo}--\ref{prop:svi} restate or reprove results discussed informally elsewhere in the paper; Propositions~\ref{prop:pooling} and~\ref{prop:fano} are new contributions.

\begin{proposition}[Validity of the ELBO]
\label{prop:elbo}
For any variational distribution $q$ in the family \eqref{eq:varfamily} absolutely continuous with respect to the prior, $\elbo(q)\le \log p(F)$ with equality iff $q=p(\cdot\mid F)$.
\end{proposition}
\begin{proof}
Standard. $\log p(F)=\elbo(q)+\KL(q\|p(\cdot\mid F))\ge\elbo(q)$ by Gibbs' inequality, with equality iff the KL is zero, i.e.\ iff $q=p(\cdot\mid F)$ almost everywhere.
\end{proof}

\begin{proposition}[Hierarchical identifiability under instrumental anchors]
\label{prop:identify-restated}
Let $U$ denote unmeasured confounders, $X_{\mathrm{prior}}$ the buffy-coat input, and $G$ the mQTL genotype. Assume (i) $G\perp U$ (Mendelian randomization), (ii) $G\not\!\perp \mu^{\mathrm{pop}}$ (relevance), (iii) $G\perp F\mid (\mu^{\mathrm{pop}},U)$ (exclusion), and (iv) the conditional distribution $p(\mu^{\mathrm{pop}}\mid G)$ is complete in the sense of \citet{newey2003instrumental} and \citet{darolles2011nonparametric}. Then the population latent $\mu^{\mathrm{pop}}$ is point-identified from the joint $(F,G,X_{\mathrm{prior}})$, and the VIB-regularized variational posterior \eqref{eq:vib} converges to the correct conditional with $\beta_{\mathrm{VIB}}\to\infty$ in the population limit.
\end{proposition}
\begin{proof}
This is the statement of Proposition~\ref{prop:identify} in Section~\ref{sec:identify}, reproduced here for completeness; see that section for the proof sketch.
\end{proof}

\begin{proposition}[Asymptotic VIB-bounded prior leakage]
\label{prop:vib}
Let $\eta_{\mathrm{leak}}(q)=I_q(\zeta;X_{\mathrm{prior}})$ where $\zeta$ is the encoder representation under $q$. Under the VIB-regularized objective $\elbo_{\mathrm{VIB}}=\elbo - \beta_{\mathrm{VIB}}\,\E_{X_{\mathrm{prior}}}[\KL(q_\phi(\zeta\mid X_{\mathrm{prior}})\|r(\zeta))]$ with $r(\zeta)=\N(0,I)$, any sequence of optima $q^*_{\beta_{\mathrm{VIB}}}$ as $\beta_{\mathrm{VIB}}\to\infty$ satisfies $\eta_{\mathrm{leak}}(q^*_{\beta_{\mathrm{VIB}}})\to 0$.
\end{proposition}
\begin{proof}
The VIB upper bound on mutual information \citep{alemi2017vib} gives
\[
\eta_{\mathrm{leak}}(q) \le \E_{X_{\mathrm{prior}}}\big[\KL(q_\phi(\zeta\mid X_{\mathrm{prior}})\|r(\zeta))\big].
\]
Under $\elbo_{\mathrm{VIB}}$, the optimum trades off reconstruction against this upper bound. As $\beta_{\mathrm{VIB}}\to\infty$, the regularized objective forces the upper bound to zero in the limit (otherwise the regularizer dominates and the objective diverges to $-\infty$), and hence $\eta_{\mathrm{leak}}\to 0$.
\end{proof}

\begin{corollary}[Finite-$\beta_{\mathrm{VIB}}$ leakage bound]
\label{cor:vib-finite}
Let $G(\beta_{\mathrm{VIB}})=\elbo(q_{\mathrm{ML}})-\elbo(q^*_{\beta_{\mathrm{VIB}}})$ be the gap between the unregularized maximum-likelihood ELBO and the VIB-regularized optimum. Then
\[
\eta_{\mathrm{leak}}(q^*_{\beta_{\mathrm{VIB}}}) \le \frac{G(\beta_{\mathrm{VIB}})}{\beta_{\mathrm{VIB}}}.
\]
\end{corollary}
\begin{proof}
By construction, $\elbo_{\mathrm{VIB}}(q^*_{\beta_{\mathrm{VIB}}}) \ge \elbo_{\mathrm{VIB}}(q_{\mathrm{ML}}) = \elbo(q_{\mathrm{ML}}) - \beta_{\mathrm{VIB}}\cdot\eta_{\mathrm{leak}}(q_{\mathrm{ML}})\ge \elbo(q_{\mathrm{ML}}) - \beta_{\mathrm{VIB}}\cdot\eta_{\mathrm{leak}}^{\max}$ where $\eta_{\mathrm{leak}}^{\max}$ is a uniform upper bound on the leakage of any feasible $q$ (which exists because mutual information is bounded by entropy of $X_{\mathrm{prior}}$). Rearranging, $\beta_{\mathrm{VIB}}\cdot\eta_{\mathrm{leak}}(q^*_{\beta_{\mathrm{VIB}}})\le \elbo(q_{\mathrm{ML}})-\elbo(q^*_{\beta_{\mathrm{VIB}}}) = G(\beta_{\mathrm{VIB}})$, giving the bound.
\end{proof}

\begin{proposition}[Conformal marginal coverage]
\label{prop:conformal-restated}
For any joint distribution of $(F,\beta^*)$ and any model $q(\beta\mid F)$, if calibration and test points are exchangeable, the conformal set $C(F_{\mathrm{new}})$ defined in Section~\ref{sec:calibration} satisfies $\Pr(\beta^*_{\mathrm{new}}\in C(F_{\mathrm{new}}))\ge 1-\alpha$.
\end{proposition}
\begin{proof}
This is the statement of Proposition~\ref{prop:conformal} in Section~\ref{sec:calibration}, reproduced here for completeness; the proof appears in that section.
\end{proof}

\begin{proposition}[Convergence of non-conjugate natural-gradient SVI]
\label{prop:svi}
Let $\lambda^{(t)}$ denote the variational natural parameters of the population layer under the modified non-conjugate update of \citet{khan2017cvi,lin2019natgrad}, applied with the moment-matching target \eqref{eq:nat-target} and Robbins-Monro step sizes $\rho_t$ satisfying $\sum\rho_t=\infty$, $\sum\rho_t^2<\infty$. Under bounded-gradient-norm regularity on the encoder and the local layer, the iterates $\lambda^{(t)}$ converge almost surely to a stationary point of the ELBO.
\end{proposition}
\begin{proof}
The bare statement of \citet{hoffman2013svi} Theorem 1 does not directly apply because the local layer $q_\phi(\eta\mid F)$ is a normalizing flow, not a member of the conjugate exponential family conditional on $\mu^{\mathrm{pop}}$. \citet{khan2017cvi,lin2019natgrad} extend natural-gradient SVI to non-conjugate models by replacing the conjugate sufficient statistics with moment-matched proxies $\hat\lambda^{\mathrm{nat}}_\ell$ computed from the encoder posterior. Under bounded-gradient-norm regularity on the encoder and the local layer, this preserves the unbiasedness of the gradient estimator (in the natural parameterization) up to vanishing bias from the moment-matching, and combined with \citet{robbins1951} stochastic-approximation conditions on $\rho_t$, the iterates converge a.s.\ to a stationary point of the regularized ELBO. The specific regularity conditions are stated as Assumptions A1--A4 of \citet{lin2019natgrad}.
\end{proof}

\begin{proposition}[Variance reduction from hierarchical pooling]
\label{prop:pooling}
Let $\hat\beta^{\mathrm{indep}}_{s,\ell}$ be the per-sample posterior mean estimator of $\beta_{s,\ell}$ from per-sample inference (no population pooling), and let $\hat\beta^{\mathrm{HAVI}}_{s,\ell}$ be the per-sample posterior mean under the HAVI-Methyl hierarchical model with population variance $\tau_0^2$ and per-sample observation precision $\tau_{\mathrm{obs}}^{-2}$. Assume the population mean $\mu^{\mathrm{pop}}_\ell$ is well-estimated, with population variance $\hat\tau_{\mathrm{pop}}^2$. Then on the logit scale,
\[
\Var(\hat\eta^{\mathrm{HAVI}}_{s,\ell}) \le \frac{\hat\tau_{\mathrm{pop}}^2 \cdot \tau_{\mathrm{obs}}^2}{\hat\tau_{\mathrm{pop}}^2 + \tau_{\mathrm{obs}}^2} < \tau_{\mathrm{obs}}^2 = \Var(\hat\eta^{\mathrm{indep}}_{s,\ell}),
\]
with strict inequality whenever $\hat\tau_{\mathrm{pop}}^2 < \infty$. The variance reduction factor is $\rho_\ell=\hat\tau_{\mathrm{pop}}^2/(\hat\tau_{\mathrm{pop}}^2+\tau_{\mathrm{obs}}^2)\in[0,1)$, with $\rho_\ell\to 0$ when the population is highly concentrated (strong pooling) and $\rho_\ell\to 1$ when the population is diffuse (weak pooling).
\end{proposition}
\begin{proof}
On the logit scale, the model is a Gaussian-Gaussian conjugate hierarchy with prior $\mu^{\mathrm{pop}}_\ell\sim\N(\mu_0,\tau_0^2)$, per-sample observation $\eta_{s,\ell}\mid\mu^{\mathrm{pop}}_\ell\sim\N(\mu^{\mathrm{pop}}_\ell+\delta_s,\tau_{\mathrm{obs}}^2)$. The per-sample independent estimator is the maximum-likelihood estimate from a single observation: $\hat\eta^{\mathrm{indep}}=\eta_{s,\ell}$ with variance $\tau_{\mathrm{obs}}^2$. The hierarchical posterior mean is the conjugate posterior $\hat\eta^{\mathrm{HAVI}}_{s,\ell}=(\hat\tau_{\mathrm{pop}}^{-2}\hat\mu^{\mathrm{pop}}_\ell + \tau_{\mathrm{obs}}^{-2}\eta_{s,\ell})/(\hat\tau_{\mathrm{pop}}^{-2}+\tau_{\mathrm{obs}}^{-2})$, with posterior variance $1/(\hat\tau_{\mathrm{pop}}^{-2}+\tau_{\mathrm{obs}}^{-2})=\hat\tau_{\mathrm{pop}}^2\tau_{\mathrm{obs}}^2/(\hat\tau_{\mathrm{pop}}^2+\tau_{\mathrm{obs}}^2)$. This is strictly less than $\tau_{\mathrm{obs}}^2$ for any finite $\hat\tau_{\mathrm{pop}}^2$, completing the bound. This is the textbook James-Stein / shrinkage-estimation result; the proposition states it in the context of the HAVI-Methyl hierarchy.
\end{proof}

\begin{proposition}[Information-theoretic recoverability lower bound]
\label{prop:fano}
Let $\hat\beta_{s,\ell}(F_{s,\ell})$ be any estimator of $\beta_{s,\ell}$ from the fragment bag $F_{s,\ell}$, and let $I(\beta_{s,\ell};F_{s,\ell})$ denote the mutual information between the latent and the fragment bag under the generative model of Section~\ref{sec:model}. Discretize $\beta_{s,\ell}$ into $M$ equally-sized bins on $[0,1]$, and let $\tilde\beta_{s,\ell}$ be the bin index. For any estimator $\hat\beta_{s,\ell}$ with bin-error probability $P_e=\Pr(\hat{\tilde\beta}\ne\tilde\beta)$,
\[
P_e \ge 1 - \frac{I(\beta_{s,\ell};F_{s,\ell}) + \log 2}{\log M}.
\]
Equivalently, $\E[(\hat\beta-\beta)^2]\ge \Omega(M^{-2})\cdot(1-P_e)\ge \Omega(M^{-2})\cdot\big[(I(\beta;F)+\log 2)/\log M\big]$ asymptotically as $M\to\infty$.
\end{proposition}
\begin{proof}
This is a direct application of Fano's inequality \citep{cover2006information}, Theorem 2.10.1, to the discretized estimation problem. Fano's inequality states that for any estimator $\hat W$ of a discrete random variable $W\in\{1,\dots,M\}$ from observation $Y$, with error probability $P_e=\Pr(\hat W\ne W)$, $H(P_e)+P_e\log(M-1)\ge H(W\mid Y)$, where $H$ denotes entropy. Rearranging, $P_e\ge (H(W\mid Y)-1)/\log M=(H(W)-I(W;Y)-1)/\log M$. For $W=\tilde\beta$ uniform on $M$ bins, $H(W)=\log M$, so $P_e\ge 1-(I(W;Y)+1)/\log M\ge 1-(I(\beta;F)+\log 2)/\log M$ where the last inequality uses $I(\tilde\beta;F)\le I(\beta;F)$ by the data-processing inequality (the binning is a deterministic function of $\beta$). The MSE bound follows because squared error is at least $(2M)^{-2}$ for any prediction outside the correct bin.
\end{proof}

\begin{remark}
Proposition~\ref{prop:fano} provides a fundamental information-theoretic lower bound on the achievable estimation error: no estimator, including HAVI-Methyl, can recover $\beta_{s,\ell}$ to better mean-square error than the bound implies given the mutual information available from the fragment bag. This frames HAVI-Methyl's empirical performance as approaching a generative-model-determined ceiling, rather than as competing only against other estimators on a relative scale. Estimating $I(\beta;F)$ from the simulator output gives a quantitative benchmark for how much room HAVI-Methyl still has to grow.
\end{remark}

% ============================================================
