\section{Discussion and Limitations}
\label{sec:discussion}

\subsection{Computational cost}
The full HAVI-Methyl architecture is more expensive per training step than a per-sample HMM because it runs a fragment-bag encoder and a posterior-head sampler. The compensating advantage is amortization: a new sample is processed by encoder forward passes rather than a full per-sample Baum-Welch refit. The released torch SVI loop fits the $S=77 \times L=782$ Liu 2024 panel in 200 mini-batch iterations and approximately 80 minutes on a single NVIDIA A10G ($\sim$80~min wall-clock; cf.\ Fig.~\ref{fig:elbo}). Wafer-scale throughput estimates remain analytical extrapolations (cf.\ Sec.~\ref{sec:benchmark}, \texttt{bench\_compute\_budget.csv} rows tagged \emph{planning estimate}) and should be re-measured on the target accelerator.

\subsection{Failure modes}
Several regimes warrant caution. Deeply hypo-methylated partially methylated domains can saturate fragmentomic signals and should be flagged from public PMD tracks before downstream clock or tissue-of-origin analysis. Centromeric and pericentromeric repeats suffer from collapsed mappability, detectable from per-locus mappability scores. Short reads below the chosen QC threshold carry degraded motif information. Low-coverage tail loci with $n^{\mathrm{frag}}<2$ should be reported as posterior-prior fallbacks rather than informative posteriors. The implementation should emit diagnostics for these cases: PMD overlap, mappability flags, read-length distributions, local coverage, posterior effective sample size, and stratum-specific calibration.

\subsection{Extensions}
Several extensions are natural follow-ups. Nanopore direct-methylation likelihoods can replace the Beta-Binomial pseudo-likelihood with the Bernoulli direct-call form. Joint inference with ichorCNA \citep{adalsteinsson2017ichor} could provide CNV-corrected methylation values for cancer cfDNA. A methylation foundation model such as CpGPT or MethylGPT can be attached downstream as an auxiliary consumer of the HAVI-Methyl posterior $\beta$ output, rather than as a direct WGS-fragment baseline.

\subsection{Clinical translation}
Clinical translation of an ML-based methylation predictor requires prospective validation, subgroup analysis, calibration monitoring, and regulatory review. The conformal wrapper supplies a distribution-free marginal coverage guarantee under exchangeability, but deployment still requires empirical auditing across age, sex, ancestry, cohort, batch, and disease strata. The mQTL anchor loss is only as representative as the genotype panels and summary statistics supporting it, so ancestry-specific or multi-ancestry anchor evaluations should be planned. Regulatory details should be verified against current guidance before final claims are made.

\subsection{Open follow-ups}
The released stack ships the full proposed method, proof sketches and conditional theory, end-to-end synthetic and real-data measurements, conformal calibration on the synthetic ablation matrix, and tissue-of-origin LOO on the Loyfer/UXM\_deconv U25 atlas. Research follow-ups that remain open are: (i) the doubly-reparameterized gradient (DReG) variance reduction for the IWAE objective, which requires \texttt{torch.func.functional\_call} encoder-parameter detachment and was not benefit-positive in the simplified surrogate; and (ii) prospective clinical validation with calibrated cfDNA assays under regulatory protocol. Three follow-ups previously listed here are closed in this revision: (a) all five App.~H validation axes are now \emph{verified} on the chromatin-aware simulator with the 3-mode mixture re-fit to real Liu 2024 fragments (App.~\ref{app:simvalidation}); (b) a true gradient-reversal adversarial head with a sample-id discriminator now replaces the context-variance proxy (Sec.~\ref{sec:identify}); (c) joint training of the variance-weighted Dirichlet tissue head inside \texttt{fit\_svi\_torch} is wired through the optional \texttt{tissue\_reference}/\texttt{tissue\_target}/\texttt{tissue\_weight} kwargs (Sec.~\ref{sec:too}).

% ============================================================
