\section{Background and Preliminaries}
\label{sec:background}

\subsection{cfDNA fragmentomics}
cfDNA fragments are not random shears: they are produced by a programmed combination of apoptotic caspase-activated DNase, DNase1L3, and other endonucleases acting on chromatinized DNA, and the resulting fragment-length distribution carries a $\sim$10.4-bp periodicity reflecting nucleosome rotational positioning \citep{snyder2016,jiang2020}. We use input features drawn from this literature. Fragment length carries a dominant signal, with the canonical $167$~bp mononucleosomal mode and secondary peaks near di- and tri-nucleosomal multiples \citep{snyder2016}. Orientation-aware fragmentation near transcription start sites motivates the orientation feature \citep{sun2019orientationaware}; end-motif differences motivate the 4-mer end-motif feature \citep{zhou2022}; and genome-wide fragmentation profiles motivate fragment-feature prediction more generally \citep{cristiano2019delfi}. FinaleMe \citep{liu2024finaleme} unified these signals into a sequence-level HMM that predicts methylation. HAVI-Methyl uses a related feature set---fragment length, end motif, sequence context, coverage, orientation, mappability, and inferred nucleosome distance---but separates the generative observation model from the amortized recognition network.

\paragraph{Observation regimes.} The manuscript distinguishes three regimes that must not be conflated. In direct methylation regimes, such as bisulfite or nanopore experiments, per-CpG methylation indicators can be treated as observations. In paired-validation or pseudo-label regimes, locus-level methylated-count summaries are outputs of a caller or classifier and should enter as pseudo-observations with an explicit over-dispersion parameter. In WGS-only deployment, fragment features are observed directly and methylation is inferred by the recognition model and calibrated against held-out truth when available. Sections~\ref{sec:model} and~\ref{sec:elbo} keep these regimes separate so that planned WGS-only experiments do not read as completed direct methylation assays.

\subsection{Variational Bayesian inference}
Given observations $X$ and latents $Z$, variational inference \citep{jordan1999,wainwright2008} approximates the posterior $p(Z\mid X)$ by a variational distribution $q_\phi(Z)$ chosen to maximize the evidence lower bound
\begin{equation}
\elbo(\phi) = \E_{q_\phi(Z)}[\log p(X,Z) - \log q_\phi(Z)] = \log p(X) - \KL(q_\phi(Z)\,\|\,p(Z\mid X)).
\label{eq:elbo-canonical}
\end{equation}
We organize the relevant prior work into three threads.

\paragraph{Classical and amortized VI.} \citet{jordan1999,wainwright2008,blei2017vi} laid the foundations of mean-field VI for graphical models. \citet{kingma2014vae} and \citet{rezende2014} introduced amortized VI through a recognition network $q_\phi(Z\mid X)$ trained by reparameterization gradients, which is the basis for the Set-Transformer-conditioned encoder used here.

\paragraph{Scalable VI.} \citet{hoffman2013svi} showed that for conditionally conjugate exponential-family models, natural-gradient ascent in $\elbo$ becomes a stochastic forward-pass update on natural parameters, enabling stochastic VI to scale to billions of observations. \citet{hoffmanblei2015ssvi} extended SVI to structured variational families. \citet{titsias2014doublystoch} introduced doubly stochastic VI for non-conjugate models, and \citet{kucukelbir2017advi} introduced ADVI as a general-purpose automatic VI engine. HAVI-Methyl uses an approximate population-layer update inspired by SVI, with the non-conjugate moment-matching caveats discussed in Sections~\ref{sec:elbo} and~\ref{sec:theory}.

\paragraph{Expressive VI.} \citet{rezende2015flows}, \citet{kingma2016iaf}, and \citet{durkan2019nsf} replaced Gaussian variational families with normalizing flows---invertible neural maps with tractable Jacobians---yielding flexible posteriors. \citet{burda2016iwae} introduced importance-weighted bounds (IWAE); \citet{rainforth2018iwaeSNR} showed that the encoder gradient signal-to-noise ratio can degrade as the IWAE sample count grows; and \citet{tucker2019dreg} introduced the doubly-reparameterized gradient estimator (DReG). \citet{alemi2017vib} developed the variational information bottleneck, which we use to limit information flow from the specified prior input rather than to claim elimination of all unmeasured confounding.

\subsection{Signal-to-component map}
Fragment length and inferred nucleosome distance primarily enter through the fragment-bag encoder and, in the full generative specification, the coverage/fragmentation terms. End motifs enter both the fragment representation and the optional Categorical likelihood. GC content and mappability enter the Negative-Binomial coverage component and the encoder context. Long-range sequence context enters through a frozen or jointly trained DNA encoder. Tissue-reference panels enter only the tissue-of-origin head and should be treated as downstream reference information, not as observed methylation truth for the WGS-only methylation task.

\subsection{Methylation foundation models, clocks, and probabilistic methylation prediction}
Recent work has begun to train transformer-style foundation models on the methylome---CpGPT \citep{cpgpt2024} and MethylGPT \citep{methylgpt2024}---which produce CpG embeddings but require methylation values as input. HAVI-Methyl complements these by producing methylation predictions \emph{from raw WGS}, after which any clock---PhenoAge, GrimAge, DunedinPACE, or PC-clocks \citep{horvath2013,levine2018phenoage,lu2019grimage,belsky2022dunedinpace,higginschen2022pcclocks}---can be applied. \citet{ying2024causality} motivates mQTL-based instrumental analyses. The closest precedents in probabilistic methylation prediction are DeepCpG \citep{angermueller2017deepcpg}, METHimpute \citep{taudt2018methimpute}, and Methyl-BERT \citep{lee2018methylbert}. These references support the positioning of HAVI-Methyl, but any external performance comparisons in the planned benchmark should be rechecked against the cited papers before final submission.

% ============================================================
