\section{Discussion and Limitations}
\label{sec:discussion}

\subsection{Computational cost}
The full HAVI-Methyl architecture is more expensive per training step than a per-sample HMM because it runs a fragment-bag encoder and a posterior-head sampler. The compensating advantage is amortization: a new sample is processed by encoder forward passes rather than a full per-sample Baum-Welch refit. The released torch SVI loop fits the $S=77 \times L=782$ Liu 2024 panel in $500$ mini-batch iterations on a single NVIDIA A10G (cf.\ Fig.~\ref{fig:elbo}; mini-batch $|B_s|=16$, $|B_\ell|=128$, $K=4$ IWAE samples). Wafer-scale throughput estimates remain analytical extrapolations (cf.\ Sec.~\ref{sec:benchmark}, \texttt{bench\_compute\_budget.csv} rows tagged \emph{planning estimate}) and should be re-measured on the target accelerator.

\subsection{Failure modes}
Several regimes warrant caution. Deeply hypo-methylated partially methylated domains can saturate fragmentomic signals and should be flagged from public PMD tracks before downstream clock or tissue-of-origin analysis. Centromeric and pericentromeric repeats suffer from collapsed mappability, detectable from per-locus mappability scores. Short reads below the chosen QC threshold carry degraded motif information. Low-coverage tail loci with $n^{\mathrm{frag}}<2$ should be reported as posterior-prior fallbacks rather than informative posteriors. The implementation should emit diagnostics for these cases: PMD overlap, mappability flags, read-length distributions, local coverage, posterior effective sample size, and stratum-specific calibration.

\subsection{Extensions}
Several extensions are natural follow-ups. Nanopore direct-methylation likelihoods can replace the Beta-Binomial pseudo-likelihood with the Bernoulli direct-call form. Joint inference with ichorCNA \citep{adalsteinsson2017ichor} could provide CNV-corrected methylation values for cancer cfDNA. A methylation foundation model such as CpGPT or MethylGPT can be attached downstream as an auxiliary consumer of the HAVI-Methyl posterior $\beta$ output, rather than as a direct WGS-fragment baseline.

\subsection{Clinical translation}
Clinical translation of an ML-based methylation predictor requires prospective validation, subgroup analysis, calibration monitoring, and regulatory review. The conformal wrapper supplies a distribution-free marginal coverage guarantee under exchangeability, but deployment still requires empirical auditing across age, sex, ancestry, cohort, batch, and disease strata. The mQTL anchor loss is only as representative as the genotype panels and summary statistics supporting it, so ancestry-specific or multi-ancestry anchor evaluations should be planned.

\paragraph{Prospective validation roadmap.} The retrospective Liu 2024 results in Sec.~\ref{sec:benchmark:real} establish per-locus posterior recovery on shallow paired cfDNA WGS / WGBS. A prospective evaluation should layer three additional cohorts on top of that baseline. (i)~\emph{Healthy adult plasma at multiple coverage depths} ($\sim$200 donors stratified by age and ancestry, with paired WGS and WGBS or EM-seq), to measure the absolute per-CpG calibration band as a function of coverage and to estimate the Mondrian conformal strata of \S\ref{sec:calibration} on a population that was not in the training set. (ii)~\emph{Pregnancy plasma} ($\sim$50 trimester-matched donors), where placenta-specific markers in the Loyfer panel give an external check on the tissue-of-origin head -- the fetal fraction predicted by HAVI-Methyl's Dirichlet head should track the standard SNP-based fetal-fraction estimate within the published assay tolerance. (iii)~\emph{Cancer cfDNA} (200--500 donors across multiple primaries with paired tumor methylation arrays or WGBS), where the methylation calls feed downstream MRD or TF-fraction analyses; the comparison metric is concordance with the array readout per case, not per CpG.

\paragraph{Calibration audits.} The conformal calibration set must be re-fit per assay platform (Illumina vs Element vs PacBio HiFi vs nanopore), per coverage tier, and per chromatin / mappability stratum, since each combination yields a different exchangeable population. We recommend the Mondrian conformal variant of Sec.~\ref{sec:calibration:cond}, with stratification by coverage decile, low-information / CpG-density bin, chromatin state, GC content, and mappability score, and that every clinical report carries the empirical coverage of its stratum alongside the posterior credible interval. Recalibration on a rolling validation window (eg quarterly) is the standard mitigation for distribution drift.

\paragraph{Regulatory framing.} Because HAVI-Methyl produces calibrated per-CpG posterior distributions rather than a deterministic diagnosis, the natural regulatory framing is as a software input to a clinical decision (FDA SaMD class I/II depending on the downstream test, or the corresponding IVDR/MDR class A/B in the EU). The relevant FDA guidances are the 2021 \emph{AI/ML SaMD Action Plan} and the 2024 \emph{Predetermined Change Control Plan} guidance for model updates. A submission package would include: model card with intended use and population, retrospective performance with bootstrap CIs and stratified subgroup analysis, conformal calibration audit on the held-out cohort, drift monitoring plan with rolling-window recalibration triggers, and an algorithm change protocol that locks the inference graph against drift between updates. The conformal wrapper's distribution-free coverage guarantee, the Sec.~\ref{sec:identify} VIB / GRL leakage controls, and the multi-seed bootstrap intervals of Sec.~\ref{sec:multiseed} are the technical artifacts most directly aligned with these reviewer asks. Regulatory specifics should still be verified against the current guidance and against jurisdiction-specific labelling rules before final claims.

\subsection{Open follow-ups}
The released stack ships the full proposed method, proof sketches and conditional theory, end-to-end synthetic and real-data measurements, conformal calibration on the synthetic ablation matrix, and tissue-of-origin LOO on the Loyfer/UXM\_deconv U25 atlas. The only research follow-up that remains open is prospective clinical validation with calibrated cfDNA assays under regulatory protocol. Four follow-ups previously listed in earlier drafts are closed in this revision: (a) all five App.~H validation axes are now \emph{verified} on the chromatin-aware simulator with the 3-mode mixture re-fit to real Liu 2024 fragments (App.~\ref{app:simvalidation}); (b) a true gradient-reversal adversarial head with a sample-id discriminator replaces the context-variance proxy (Sec.~\ref{sec:identify}); (c) joint training of the variance-weighted Dirichlet tissue head inside \texttt{fit\_svi\_torch} is wired through the optional \texttt{tissue\_reference}/\texttt{tissue\_target}/\texttt{tissue\_weight} kwargs (Sec.~\ref{sec:too}); (d) the proper Tucker-2019 doubly-reparameterised IWAE estimator (DReG) is wired through \texttt{iwae\_dreg} with the encoder parameters detached in $\log q_\phi$ at evaluation time (Sec.~\ref{sec:elbo}).

% ============================================================
