\section{Probabilistic Model}
\label{sec:model}

Let $s\in\{1,\dots,S\}$ index samples and $\ell\in\{1,\dots,L\}$ index CpG loci or windows. Let $F_{s,\ell}=\{f_{s,\ell,j}\}_{j=1}^{n^{\mathrm{frag}}_{s,\ell}}$ denote the observed fragment-feature bag used by the recognition network. Each $f_{s,\ell,j}\in\R^d$ encodes fragment length, 5'- and 3'-end 4-mer motifs, GC content, orientation, mappability, and inferred distance to the nearest nucleosome center. We write $\mathcal{O}_{s,\ell}$ for the observed generative modalities at a locus: direct methylation calls $\mathcal{Y}_{s,\ell}$ when available, pseudo-counts $(n^m_{s,\ell},n^{\mathrm{cpg}}_{s,\ell})$ when only aggregate labels are available, end motifs $\mathcal{M}_{s,\ell}$, and fragment coverage $n^{\mathrm{frag}}_{s,\ell}$. Thus $F_{s,\ell}$ is the recognition input, while $\mathcal{O}_{s,\ell}$ denotes the likelihood-bearing observations.

\subsection{Three-level hierarchy}
We place a three-level hierarchy directly on the logit-methylation latent $\eta_{s,\ell}=\logit(\beta_{s,\ell})\in\R$, parameterizing the population locus prior on the same logit scale:
\begin{align}
\mu^{\mathrm{pop}}_\ell &\sim \N(\mu_0, \tau_0^2), \quad \ell=1,\dots,L, \label{eq:pop}\\
\delta_s &\sim \N(0, \sigma_\delta^2), \quad s=1,\dots,S, \label{eq:samp}\\
\eta_{s,\ell}\mid \mu^{\mathrm{pop}}_\ell, \delta_s &\sim \N\big(\mu^{\mathrm{pop}}_\ell+\delta_s,\,\sigma_\eta^2\big),\label{eq:cell}\\
\beta_{s,\ell} &= \sigm(\eta_{s,\ell}),\qquad
\beta^{\mathrm{pop}}_\ell = \sigm(\mu^{\mathrm{pop}}_\ell).
\end{align}
The latent $\mu^{\mathrm{pop}}_\ell$ is the logit-scale population mean for locus $\ell$; $\beta^{\mathrm{pop}}_\ell\in(0,1)$ is the corresponding methylation level reported to the user. The Gaussian prior on the logit scale gives heavier tails near $\{0,1\}$ than a Beta prior would, which is useful for hyper- and hypo-methylated loci.

\paragraph{Identifiability constraint.} The additive form $\eta_{s,\ell}=\mu^{\mathrm{pop}}_\ell+\delta_s+\mathrm{noise}$ is symmetric under the shift $\mu^{\mathrm{pop}}_\ell\mapsto\mu^{\mathrm{pop}}_\ell+c$, $\delta_s\mapsto\delta_s-c$, which leaves every $\eta_{s,\ell}$ invariant. We break this symmetry by imposing a sum-to-zero constraint on the per-sample shifts,
\begin{equation}
\sum_{s=1}^{S}\delta_s=0,
\label{eq:sumzero}
\end{equation}
implemented in the variational posterior by a global or running recentering correction on the sample-shift means, not by a mini-batch-only correction. The mean-zero prior on $\delta_s$ provides additional soft identifiability through the prior mean.

\begin{assumption}[Observation and hierarchy conventions]
\label{ass:obs}
Throughout the likelihood derivations, conditional independence is assumed only after conditioning on the latent $\eta_{s,\ell}$, sequence context, mappability/GC covariates, and any stated sample-level covariates. Missing observation modalities are omitted from the likelihood rather than imputed as observed. When WGS-only fragment features are used without direct methylation calls, aggregate methylation counts are treated as pseudo-observations only in training or validation settings where such labels are supplied by a separate caller or assay.
\end{assumption}

\subsection{Likelihoods}
Conditional on $\beta_{s,\ell}$ and the relevant covariates, the observed generative modalities factorize as follows.

\paragraph{Per-fragment-CpG Bernoulli likelihood (direct-call form).} For each CpG $k$ inside fragment $j$ overlapping locus $\ell$ in sample $s$, a direct methylation indicator $y_{s,\ell,j,k}\in\{0,1\}$ is drawn Bernoulli with probability $\beta_{s,\ell}$:
\begin{equation}
y_{s,\ell,j,k}\mid \beta_{s,\ell}\sim \mathrm{Bernoulli}(\beta_{s,\ell}).
\label{eq:bern}
\end{equation}
This form is appropriate when methylation is directly measured or validated. Fragment features $f_{s,\ell,j}$ enter the encoder of the variational posterior (Section~\ref{sec:varfamily}); unless a specific feature likelihood is stated, they should not be silently treated as if they were direct methylation labels.

\paragraph{Beta-Binomial pseudo-likelihood (aggregate form).} When per-CpG direct calls are unavailable and only locus-level methylated-count summaries are observed or generated by a classifier, we aggregate to a Beta-Binomial pseudo-likelihood:
\begin{equation}
n^m_{s,\ell}\mid \beta_{s,\ell}, n^{\mathrm{cpg}}_{s,\ell} \sim \BB\big(n^{\mathrm{cpg}}_{s,\ell},\, \kappa\beta_{s,\ell},\,\kappa(1-\beta_{s,\ell})\big),
\label{eq:bb}
\end{equation}
where $n^{\mathrm{cpg}}_{s,\ell}=\sum_j n^{\mathrm{cpg}}_{s,\ell,j}$ is the number of CpG trials and $n^m_{s,\ell}$ is the number called methylated. We call this a pseudo-likelihood because, in WGS-only settings, these labels are outputs of a separate validation or classifier pipeline rather than direct WGS observations. The concentration $\kappa>0$ shrinks the effective sample size to absorb classifier uncertainty and over-dispersion.

\paragraph{Categorical end-motif likelihood, sequence-conditioned.} The 4-mer end motif $m_{s,\ell,j}\in\{1,\dots,256\}$ is dominantly determined by genomic sequence context at the cut site, with methylation acting as a secondary modifier of cut-site preference \citep{zhou2022}. We therefore condition the categorical on both methylation and a frozen or learned sequence-context embedding $c^{\mathrm{seq}}_\ell$:
\begin{equation}
m_{s,\ell,j}\mid \beta_{s,\ell}, c^{\mathrm{seq}}_\ell \sim \Cat\!\big(\bpi(\beta_{s,\ell}, c^{\mathrm{seq}}_\ell)\big),\quad
\bpi(\beta, c) = \mathrm{softmax}\!\big(W_m\,h_\theta(\beta, c) + b_m\big).
\end{equation}
Without sequence conditioning, the model would be forced to predict end motifs from methylation alone, which would be an identifiability problem because local sequence is the primary determinant of the motif.

\paragraph{Negative-Binomial coverage.} Coverage $n^{\mathrm{frag}}_{s,\ell}$ at locus $\ell$ depends on GC content, mappability, and methylation through fragmentation bias. We use a mean-dispersion parameterization
\begin{equation}
n^{\mathrm{frag}}_{s,\ell}\mid \beta_{s,\ell}, \mathrm{GC}_\ell, \mathrm{map}_\ell \sim \NB\!\big(r_\ell,\mu_{s,\ell}^{\mathrm{cov}}\big),
\label{eq:nb}
\end{equation}
where $\E[n^{\mathrm{frag}}_{s,\ell}]=\mu_{s,\ell}^{\mathrm{cov}}$, $\Var(n^{\mathrm{frag}}_{s,\ell})=\mu_{s,\ell}^{\mathrm{cov}}+(\mu_{s,\ell}^{\mathrm{cov}})^2/r_\ell$, and
\begin{equation}
\log\mu_{s,\ell}^{\mathrm{cov}}=a_0+a_\beta\beta_{s,\ell}+a_{\mathrm{GC}}\mathrm{GC}_\ell+a_{\mathrm{map}}\mathrm{map}_\ell.
\end{equation}
The dispersion $r_\ell>0$ may be shared, locus-specific, or regressed on locus covariates. Although coverage is weaker than direct read evidence, it can help disambiguate regions where fragmentation is methylation-dependent.

\subsection{Joint distribution}
The full joint, in the direct-call Bernoulli form, is
\begin{align}
p(\bm{\mu}^{\mathrm{pop}},\bdelta,\bm\eta,\mathcal{O})
=\;& \prod_\ell p(\mu^{\mathrm{pop}}_\ell)\,
     \prod_s p(\delta_s)\,
     \prod_{s,\ell} p(\eta_{s,\ell}\mid\mu^{\mathrm{pop}}_\ell,\delta_s)\nonumber\\
&\times\prod_{s,\ell,j,k} p(y_{s,\ell,j,k}\mid\beta_{s,\ell})\,
   \prod_{s,\ell,j} p(m_{s,\ell,j}\mid\beta_{s,\ell}, c^{\mathrm{seq}}_\ell)\,
   \prod_{s,\ell} p(n^{\mathrm{frag}}_{s,\ell}\mid\beta_{s,\ell}, \mathrm{GC}_\ell,\mathrm{map}_\ell).
\label{eq:joint}
\end{align}
The Beta-Binomial aggregate form is obtained by replacing the inner Bernoulli product with the single Beta-Binomial term \eqref{eq:bb}. If only a subset of modalities is available, the corresponding products are omitted.

\paragraph{Choice of likelihood form.} We use the Beta-Binomial pseudo-likelihood \eqref{eq:bb} when per-CpG calls are themselves the output of a separate classifier or validation assay whose uncertainty we wish to absorb, and the Bernoulli form \eqref{eq:bern} when per-CpG calls are direct. Both forms yield the same hierarchy on $\eta_{s,\ell}$ and differ only in the reconstruction term of the ELBO.

\subsection{Prior choice}
The hyperparameters $(\mu_0,\tau_0,\sigma_\delta,\sigma_\eta,\kappa)$ admit principled defaults but should be set from data when possible. We use $\mu_0=0$ and $\tau_0=2.0$ as a broad logit-scale prior covering most of the $\beta\in(0,1)$ range. The concentration $\kappa$ controls the effective sample size of the Beta-Binomial relative to a Binomial; an empirical-Bayes setting can be fitted from replicate variance when replicate methylation measurements are available. The standard deviations $\sigma_\delta$ and $\sigma_\eta$ can be estimated by empirical Bayes from a held-out methylation-array dataset by fitting a one-way Gaussian random-effects model on logit-$\beta$. Full-model defaults and simplified-harness values are separated in Appendix~\ref{app:hparams}.


\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  node distance=1.4cm,
  every node/.style={font=\small},
  latentnode/.style={draw,circle,minimum size=8mm,align=center},
  obsnode/.style={draw,rectangle,rounded corners,minimum height=7mm,align=center},
  arr/.style={-Latex}
]
  \node[latentnode] (mu0) {$\mu_0,\tau_0$};
  \node[latentnode, right=2.0cm of mu0] (mupop) {$\mu^{\mathrm{pop}}_\ell$};
  \node[latentnode, right=2.0cm of mupop] (eta) {$\eta_{s,\ell}$};
  \node[latentnode, above=1.0cm of eta] (delta) {$\delta_s$};
  \node[obsnode, right=2.0cm of eta] (obs) {$\mathcal{O}_{s,\ell}$};
  \draw[arr] (mu0) -- (mupop);
  \draw[arr] (mupop) -- (eta);
  \draw[arr] (delta) -- (eta);
  \draw[arr] (eta) -- (obs);
  \node[draw,dashed,fit=(mupop),inner sep=8pt,label=below:$L$] {};
  \node[draw,dashed,fit=(delta),inner sep=8pt,label=above:$S$] {};
  \node[draw,dashed,fit=(eta)(obs),inner sep=10pt,label=below:$S\times L$] {};
\end{tikzpicture}
\caption{Plate diagram of the HAVI-Methyl hierarchy. Population locus latents $\mu^{\mathrm{pop}}_\ell$ and per-sample shifts $\delta_s$ combine into $\eta_{s,\ell}=\logit\beta_{s,\ell}$, which generates the available observation modalities $\mathcal{O}_{s,\ell}$ through the selected Bernoulli, Beta-Binomial, Categorical, and Negative-Binomial terms. Fragment bags $F_{s,\ell}$ are also used by the amortized recognition model.}
\label{fig:plate}
\end{figure}


% ============================================================
